# (PART) Characters and Strings in R {-}


# Introductory Appetizer {#intro}


To give you an idea of some of the things we can do in R with string 
processing, let's play a bit with a simple example. 


## A Toy Example

For this crash informal introduction, we'll use the data frame `USArrests` 
that already comes with R. Use the function `head()` to get a peek of 
the data:

```{r USArrests}
# take a peek of USArrests
head(USArrests)
```

The labels on the rows such as `Alabama` or `Alaska` are displayed 
strings. Likewise, the labels of the columns---`Murder`, `Assault`,
`UrbanPop` and `Rape`---are also strings.


### Abbreviating strings

Suppose we want to abbreviate the names of the States. Furthermore, suppose we 
want to abbreviate the names using the first four characters of each name. One 
way to do that is by using the function `substr()` which substrings a 
character vector. We just need to indicate the `start=1` and `stop=4`
positions:

```{r USArrests_strsplit}
# names of states
states <- rownames(USArrests)

# substr
substr(x = states, start = 1, stop = 4)
```

This may not be the best solution. Note that there are four states with the same 
abbreviation `"New "` (New Hampshire, New Jersey, New Mexico, New York). 
Likewise, North Carolina and North Dakota share the same name `"Nort"`. 
In turn, South Carolina and South Dakota got the same abbreviation `"Sout"`.

A better way to abbreviate the names of the states can be performed by using 
the function `abbreviate()` like so:

```{r USArrests_abbreviate}
# abbreviate state names
states2 <- abbreviate(states)

# remove vector names (for convenience)
names(states2) <- NULL
states2
```

If we decide to try an abbreviation with five letters we just simply change the 
argument `minlength = 5`

```{r USArrests_abbreviate_ex2}
# abbreviate state names with 5 letters
abbreviate(states, minlength = 5)
```



### Getting the longest name

Now let's imagine that we need to find the longest name. This implies that we need to count the number of letters in each name. The function `nchar()` comes handy for that purpose. Here's how we could do it:

```{r USArrests_longest}
# size (in characters) of each name
state_chars = nchar(states)
state_chars

# longest name
states[which(state_chars == max(state_chars))]
```



### Selecting States

To make things more interesting, let's assume that we wish to select those states containing the letter `"k"`. How can we do that? Very simple, we just need to use the function `grep()` for working with regular expressions. Simply indicate the `pattern = "k"` as follows:

```{r USArrests_k}
# get states names with 'k'
grep(pattern = "k", x = states, value = TRUE)
```

Instead of grabbing those names containing `"k"`, say we wish to select those states containing the letter `"w"`. Again, this can be done with `grep()`:

```{r USArrests_w}
# get states names with 'w'
grep(pattern = "w", x = states, value = TRUE)
```

Notice that we only selected those states with lowercase `"w"`. But what 
about those states with uppercase `"W"`? There are several options to find 
a solution for this question. One option is to specify the searched pattern as 
a character class `"[wW]"`:

```{r USArrests_wW}
# get states names with 'w' or 'W'
grep(pattern = "[wW]", x = states, value = TRUE)
```

Another solution is to first convert the state names to lower case, and then 
look for the character `"w"`, like so:

```{r USArrests_tolower_w}
# get states names with 'w'
grep(pattern = "w", x = tolower(states), value = TRUE)
```

Alternatively, instead of converting the state names to lower case we could do 
the opposite (convert to upper case), and then look for the character 
`"W"`, like so:

```{r USArrests_toupper_W}
# get states names with 'W'
grep(pattern = "W", x = toupper(states), value = TRUE)
```

A third solution involves specifying the argument 
`ignore.case=TRUE` inside `grep()`:

```{r USArrests_w_case}
# get states names with 'w'
grep(pattern = "w", x = states, value = TRUE, ignore.case = TRUE)
```



### Some computations

Besides manipulating strings and performing pattern matching operations, we can 
also do some computations. For instance, we could ask for the distribution of 
the State names' length. To find the answer we can use `nchar()`. 
Furthermore, we can plot a histogram of such distribution:

```{r USArrests_hist, out.width='65%', fig.align='center', tidy=FALSE}
summary(nchar(states))

# histogram
hist(nchar(states), las = 1, col = "gray80", main = "Histogram", 
     xlab = "number of characters in US State names")
```

Let's ask a more interesting question. What is the distribution of the vowels 
in the names of the States? For instance, let's start with the number of 
`a`'s in each name. There's a very useful function for this purpose: 
`regexpr()`. We can use `regexpr()` to get the number of times that a 
searched pattern is found in a character vector. When there is no match, we get 
a value -1.

```{r USArrests_vowel_a_ex1}
# position of a's
positions_a <- gregexpr(pattern="a", text=states, ignore.case = TRUE)

# how many a's?
num_a <- sapply(positions_a, function(x) ifelse(x[1]>0, length(x), 0))
num_a
```

If you inspect `positions_a` you'll see that it contains some negative 
numbers `-1`. This means there are no letters `a` in that name. To 
get the number of occurrences of `a`'s we are taking a shortcut with 
`sapply()`. 

The same operation can be performed by using the function 
`str_count()` from the package `"stringr"`.

```{r USArrests_vowel_a_ex2, message=FALSE}
# load stringr (remember to install it first)
library(stringr)

# total number of a's
str_count(states, "a")
```

Notice that we are only getting the number of `a`'s in lower case. Since 
`str_count()` does not contain the argument `ignore.case`, we need 
to transform all letters to lower case, and then count the number of `a`'s 
like this:

```{r USArrests_vowel_a_ex3}
# total number of a's
str_count(tolower(states), "a")
```

Once we know how to do it for one vowel, we can do the same for all the vowels:

```{r USArrests_vowels, out.width='60%', fig.align='center', tidy=FALSE}
# calculate number of vowels in each name
vowels <- c("a", "e", "i", "o", "u")
num_vowels <- vector(mode = "integer", length = 5)

for (j in seq_along(vowels)) {
  num_aux <- str_count(tolower(states), vowels[j])
  num_vowels[j] <- sum(num_aux)
}

# sort them in decreasing order
names(num_vowels) <- vowels
sort(num_vowels, decreasing = TRUE)

# barplot
barplot(num_vowels, main = "Number of vowels in USA States names", 
        border = NA, xlim = c(0, 80), las = 1, horiz = TRUE)
```

