# Boundaries {#boundaries}

__Contributor: Chitra Venkatesh__
s
Boundaries are metacharacters too and match based on what preceeds or follows current position. Let's understand this through some examples after reviewing common boundaries.

| Boundary | Description | Example    |
|:---------|:------------|:-----------|
| `\\b`    | Matches a word boundary, i.e., when a side is not `[A-z0-9_]` | `\\bHi \\bHi\\b` |
| `\\B`    | Matches when not a word boundary, i.e., when a side is `[A-z0-9_]` | `\\BHi` |

In the following example, we would try these cases:

- Case 1: Extract words that exactly match with `hi`

- Case 2: Extract words that contain `hi` in between characters

- Case 3: Extract words that start with `hi`

```{r}
some_text <- c("Hi there, its high time we learn shiny apps")
```

**Case 1:** Extract words that exactly match with `hi`.
Note the use of `ignore_case` argument to regex()

```{r}
str_extract_all(some_text, regex("\\bhi\\b", ignore_case = TRUE))
```


**Case 2:** Extract words that contain `hi` in between characters

- We include `[A-z]*` on either side of `\\Bhi\\B` to match the entire word,
otherwise the pattern `\\Bhi\\B` will only match `hi` from `shiny`. 

- We used `[A-z]*` instead of `[A-z]+` to specifically showcase that no other 
`hi` got matched as `*` denotes 0 or any.

```{r}
str_extract_all(some_text, regex("[A-z]*\\Bhi\\B[A-z]*", ignore_case = TRUE))
```

**Case 3:** Extract words that start with `hi`

Note that `shiny` is not matched since it starts with `s` and not `hi`.

```{r}
str_extract_all(some_text, regex("\\bhi[A-z]*", ignore_case = TRUE))
```

