# Character Sets

In this chapter we will talk about character sets. You will learn about 
a couple of more metacharacters, the opening and closing brackets `[ ]`, that 
will help you define a character set.

These square brackets indicate a character set which will match any one of 
the various characters that are inside the set. Keep in mind that a character 
set will match only one character. The order of the character inside the set 
does not matter; what matter is just the presence of the characters inside the
brackets. So for example if you have a set defined by `"[AEIOU]"`, that will 
match any one upper case vowel. 


## Defining character sets

Consider the following pattern that includes a character set: `"p[aeiou]n"`, 
and a vector with the words _pan_, _pen_, and _pin_:

```{r results='hide'}
pns <- c('pan', 'pen', 'pin', 'pon', 'pun')

str_view(pns, "p[aeiou]n")
```

The set `"p[aeiou]n"` matches all elements in `pns`. Now let's use the same 
set with another vector `pnx`:

```{r results='hide'}
pnx <- c('pan', 'pen', 'pin', 'p0n', 'p.n', 'p1n', 'paun')

str_view(pnx, "p[aeiou]n")
``` 

As you can tell, this time only the first three elements in `pnx` are matched.
Notice also that _paun_ is not matched. This is because the character set
matches only one character, either _a_ or _u_ but not _au_.

If you are interested in matching all capital letters, you can defined a set
formed as:

```
[ABCDEFGHIJKLMNOPQRSTUVWXYZ]
```

Likewise, you can define a set with only lower case letters:

```
[abcdefghijklmnopqrstuvwxyz]
```

If you are interested in matching any digit, you can also specify a character
set like this:

```
[0123456789]
```



## Character ranges

The previous examples that show character sets containing all the capital 
letters or all lower case letters are very convenient but require a lot typing.
Character ranges are going to help you solve that problem, by giving you a 
convenient shortcut by using the dash metacharacter `"-"` to indicate a range 
of characters. The use of the dash inside a character set represents all 
characters that are between a starting character and an ending character. 

Let's see how you can reexpress the examples in the previous section as
character ranges. The set of all digits can be expressed as a character range
using the following pattern:

```
[0-9]
```

Likewise, the set of all lower case letters _abcd...xyz_ is compactly 
represented with the character range:

```
[a-z]
```

And the character set of all upper case letters _ABD...XYZ_ is formed by

```
[A-Z]
```

It is important to note that the dash is only a metacharacter when it's inside 
a character set; outside the character set it's just a literal dash.

