# Metacharacters {#metacharacters}

The next topic that you should learn about regular expressions has to do with 
__metacharacters__. As you just learned, the most basic type of regular 
expressions are the literal characters which are characters that match 
themselves. However, not all characters match themselves. Any character that 
is not a literal character is a metacharacter. 


## About Metacharacters

Metacharacter are characters that have a special meaning and they allow you to transform literal characters in very powerful ways.

Below is the list of metacharacters in _Extended Regular Expressions_ (EREs):

```
.   \   |   (   )   [   ]   {   }   $   -    ^   *   +   ?
```

- the dot `.`
- the backslash `\`
- the bar `|`
- left or opening parenthesis `(`
- right or closing parenthesis `)`
- left or opening bracket `[`
- right or closing bracket `]`
- left or opening brace `{`
- right or closing brace `}`
- the dollar sign `$`
- the dash, hyphen or minus sign `-`
- the caret or hat `^`
- the star or asterisk `*`
- the plus sign `+`
- the question mark `?`

We're going to be working with these characters throughout the rest of the book. 
Simply put, everything else that you need to know about regular expressions 
besides literal characters is how these metacharacters work. The good news is 
that there are only a few metacharacters to learn. The bad news is 
that some metacharacters can have more than one meaning. And learning those 
meanings definitely takes time and requires hours of practice. The meaning of 
the metacharacters greatly depend on the context in which you use them,
how you use them, and where you use them. If it wasn't enough complication, 
it is also the metacharacters that have variation between the different 
regex engines. 
 


## The Wild Metacharacter

The first metacharacter you should learn about is the dot or period `"."`, 
better known as the __wild__ metacharacter. This metacharacter is used to
match __ANY__ character except for a new line.

For example, consider the pattern `"p.n"`, that is, _p wildcard n_. This
pattern will match _pan_, _pen_, and _pin_, but it will not match _prun_ 
or _plan_. The dot only matches one single character. 

Let's see another example using the vector `c("not", "note", "knot", "nut")` 
and the pattern `"n.t"`

```{r eval = FALSE}
not <- c("not", "note", "knot", "nut")

str_view(not, "n.t")
```

the pattern `"n.t"` matches _not_ in the first three elements, and _nut_
in the last element.

If you specify a pattern `"no."`, then just the first three elements
in `not` will be matched. 

```{r eval = FALSE}
str_view(not, "no.")
```

And if you define a pattern `"kn."`, then only the third element is matched.

```{r eval = FALSE}
str_view(not, "kn.")
```

The wild metacharacter is probably the most used metacharacter, and it is 
also the most abused one, being the source of many mistakes. Here is a basic
example with the regular expression formed by `"5.00"`. If you think that this 
pattern will match five with two decimal places after it, you will be 
surprised to find out that it not only matches _5.00_ but also _5100_ and _5-00_. 
Why? Because `"."` is the metacharacter that matches absolutely anything.
You will learn how to fix this mistake in the next section, but it illustrates 
an important fact about regular expressions: the challenge consists of matching 
what you want, but also in matching only what you want. You don't want to 
specify a pattern that is overly permissive. You want to find the thing you're 
looking for, but only that thing. 



## Escaping metacharacters

What if you just want to match the character dot? For example, say you 
have the following vector:

```{r}
fives <- c("5.00", "5100", "5-00", "5 00")
```

If you try the pattern `"5.00"`, it will match all of the elements in `fives`.

```{r eval = FALSE}
str_view(fives, "5.00")
```

To actually match the dot character, what you need to do is __escape__ the
metacharacter. In most languages, the way to escape a metacharacter is by 
adding a backslash character in front of the metacharacter: `"\."`. 
When you use a backslash in front of a metacharacter you are "escaping" the
character, this means that the character no longer has a special meaning, 
and it will match itself.

However, R is a bit different. Instead of using a backslash you have to use
two backslashes: `"5\\.00"`. This is because the backslash `"\"`, which is
another metacharacter, has a special meaning in R. Therefore, to match just
the element _5.00_ in `fives` in R, you do it like so:

```{r eval = FALSE}
str_view(fives, "5\\.00")
```
