# Quantifiers {#quantifiers}

Quantifiers quantify the number of instances of a character, group or character
class, denoted as `P` in the table below. Your quantifier should be placed after 
the character/group/character class that is being quantified. We will also 
explain what groups are in this context.


| Quantifier | Description                    |
|:-----------|:-------------------------------|
| `P*`       | 0 or more instances of P       |
| `P+`       | 1 or more instances of P       |
| `P?`       | 0 or 1 instance of P           |
| `P{m}`     | Exactly m instances of P       |
| `P{m,}`    | At least m instances of P      |
| `P{m,n}`   | Between m and n instances of P |

In the following example, let us try to extract all those names that contain 
more than 4 characters and less than 7 characters.

```{r}
student_names <- c("Lee", "Carol", "Sameer", "Luca", "Rajan", "George Jr.")

str_extract(student_names, regex("^[A-z]{5,7}$"))
```

In the above example, we used anchors `^` and `$` to indicate an exact match. 
In absence of which a substring of `George Jr.` also gets displayed. 

Let's try to detect names of those individuals with one or more `e` or `u`.

```{r}
str_detect(student_names, regex("[eu]+"))
```

In the last example, if we want to extract names that contain `e` or `u` we 
could follow this simple implementation . Points to note here:

- Character set `[eu]` could appear 1 or more times so we use quantifier `+`.

- `.*` matches 0 or any number of characters where `.` is a wildcard dot and 
`*` represents the quantifier 0 or many 

- Pattern `.*[eu]+.*` looks for 1 or more numbers of `[eu]` that can be 
preceeded/followed by any number of other characters.

```{r}
student_names <- c("Lee", "Carol", "Sameer", "Luca", "Rajan", "George Jr.")

str_extract(student_names, regex(".*[eu]+.*"))
```



### What do groups mean in Regex?

We visited character classes in one of the sections. For situations where we 
would like to group character classes or regex pattern before using a quantifier, 
we indicate grouping using paranthesis.

Consider an example where we would like to extract only strings with two names
separated by a whitespace. For illustrative purpose, the strings end with a 
whitespace.

```{r}
student_names <- c(
  "Lee Zhang ", 
  "Carol Roberts ", 
  "Sameer ", 
  "Luca ", 
  "Rajan ", 
  "George Smith ")

str_extract(student_names, regex("([A-z]+[ ]){2}"))
```

We could also use pre-built class `[:alpha:]` in the above example.

```{r}
str_extract(student_names, regex("([:alpha:]+[ ]){2}"))
```
