# (PART) Regex {-}

# Regular Expressions {#regex1}

## Introduction

So far you have learned some basic and intermediate functions for handling and 
working with text in R. These are very useful functions and they allow you 
to do many interesting things. However, if you truly want to unleash the power 
of strings manipulation, you need to take things to the next level and learn 
about _regular expressions_.


## What are Regular Expressions?

The name "Regular Expression" does not say much. However, regular 
expressions are all about text. Think about how much text is all around 
you in our modern digital world: email, text messages, news articles, blogs, 
computer code, contacts in your address book---all these things are text. 
Regular expressions are a tool that allows us to work with these text by 
describing text patterns.

A __regular expression__ is a special text string for describing a certain 
amount of text. This "certain amount of text" receives the formal name of 
_pattern_. In other words, a regular expression is a set of symbols that 
describes a text pattern. More formally we say that a regular expression is a 
_pattern that describes a set of strings_. In addition to this first 
meaning, the term regular expression can also be used in a slightly different 
but related way: as the formal language of these symbols that needs 
to be interpreted by a regular expression processor.
Because the term "regular expression" is rather long, most people 
use the word __regex__ as a shortcut term. And you will even find the 
plural _regexes_.

It is also worth noting what regular expressions are not. They're not a 
programming language. They may look like some sort of programming language
because they are a formal language with a defined set of rules that gets a 
computer to do what we want it to do. However, there are no variables in regex
and you can't do computations like adding 1 + 1. 


### What are Regular Expressions used for?

We use regular expressions to work with text. Some of its common uses involve
testing if a phone number has the correct number of digits, if a date follows a 
specifc format (e.g. mm/dd/yy), if an email address is in a valid format, or if 
a password has numbers and special characters. You could also use regular 
expressions to search a document for _gray_ spelt either as "gray" or 
"grey". You could search a document and replace all occurrences of Will, Bill, 
or "W." with William. Or you could count the number of times in a document 
that the word "analysis" is immediately preceded by the words "data", 
"computer" or "statistical" only in those cases. You could use it to 
convert a comma-delimited file into a tab-delimited file or to find duplicate 
words in a text. 

In each of these cases, you are going to use a regular expression to write up a 
description of what you are looking for using symbols. In the case of a phone 
number, that pattern might be three digits followed by a dash, followed by three 
digits and another dash, followed by four digits. Once you have defined a 
pattern then the regex processor will use our description to return matching 
results, or in the case of the test, to return true or false for whether or not 
it matched.


### A word of caution about regex

If you have never used regular expressions before, their syntax may seem a bit
scary and cryptic. You will see strings formed by a bunch of letters, digits, 
and other punctuation symbols combined in seemingly nonsensical ways. As with
any other topic that has to do with programming and data analysis, learning
the principles of regex and becoming fluent in defining regex patterns takes
time and requires a lot of practice. The more you use them, the better you
will become at defining more complex patterns and getting the most out of them.

Regular Expressions is a wide topic and there are books entirely dedicated to
this subject. The material offered in this book is not extensive and there are
many subtopics that I don't cover here. Despite the initial barriers that you
may encounter when entering the regex world, the pain and frustration of 
learning this tool will payoff in your data science career.


### Regular Expressions in R

Tools for working with regular expressions can be found in virtually all 
scripting languages (e.g. Perl, Python, Java, Ruby, etc). R has some 
functions for working with regular expressions but it does not provide 
the wide range of capabilities that other scripting languages do. Nevertheless, 
they can take us very far with some workarounds (and a bit of patience).

One of the best tools you must have in your toolkit is the R package `"stringr"`
(by Hadley Wickham). It provides functions that have similar behavior to 
those of the base distribution in R. But it also provides many more facilities
for working with regular expressions.


## Regex Basics

The main purpose of working with regular expressions is to describe patterns 
that are used to match against text strings. Simply put, working with regular 
expressions is nothing more than __pattern matching__. The result of a match 
is either successful or not.

The simplest version of pattern matching is to search for one occurrence (or 
all occurrences) of some specific characters in a string. For example, we might 
want to search for the word `"programming"` in a large text document, or we 
might want to search for all occurrences of the string `"apply"` in a series 
of files containing R scripts.

Typically, regular expression patterns consist of a combination of alphanumeric 
characters as well as special characters. A regex pattern can be as simple as a 
single character, or it can be formed by several characters with a more complex 
structure. In all cases we construct regular expressions much in the same form 
in which we construct arithmetic expressions, by using various operators to 
combine smaller expressions. 


## Matching Literal Characters

We're going to start with the simplest match of all: a literal character.
A literal character match is one in which a given character such as the letter
`"A"` matches the letter _A_. This type of match is the most basic 
type of regular expression operation: just matching plain text.

The following examples are extremely basic but they will help you get a 
good understanding of regex.

Consider the following text stored in a character vector `this_book`:

```{r book}
this_book <- 'This book is mine'
```

The first regular expression we are going to work with is `"book"`. 
This pattern is formed by a letter _b_, followed by a letter _o_, followed by 
another letter _o_, followed by a letter _k_. As you may guess, this pattern 
matches the word _book_ in the character vector `this_book`. 
To have a visual representation of the actual pattern that is matched, you
should use the function `str_view()` from the package `"stringr"`:

```{r}
str_view(this_book, 'book')
```

As you can tell, the pattern `"book"` doesn't match the entire content in
the vector `this_book`; it just matches those four letters.

It may seem really simple but there are a couple of details to be highlighted.
The first is that regex searches are case sensitive by default. This means 
that the pattern `"Book"` would not match _book_ in ``this_book``.

```{r}
str_view("This Book is mine.", 'book')
```

You can change the matching task so that it is case insensitive but we will
talk about it later.

Let's add more text to `this_book`:

```{r book_online}
this_book <- 'This book is mine. I wrote this book with bookdown.'
```

Let's use `str_view()` to see what pieces of text are matched in `this_book` 

```{r grep_book}
str_view(this_book, "book")
```

As you can tell, only the first occurrence of _book_ was matched. This is a 
minor but important detail and we will come back to this behavior of regular
expressions.

All the letters and digits in the English alphabet are considered literal 
characters. They are called _literal_ because they match themselves.

```{r}
str_view <- c("I had 3 quesadillas for lunch", "3")
```

Here is another example:

```{r}
transport <- c("car", "bike", "boat", "airplane")
```

The first pattern to test is the letter `"a"`:

```{r eval = FALSE}
str_view(transport, "a")
```

When you execute the previous command, you should be able to see that the 
letter `"a"` is highlighted in the words _car_, _boat_ and _airplane_.




